---
---

---
title: "Visualization with R workshop"
author: "Albina Gibadullina and Maria Laricheva"
date: "October 8, 2021"
output:
  html_document:
    theme: yeti
    toc: yes
    toc_float: yes
---

# Installing and loading packages

Install R packages
```{r}
install.packages("tidyverse")
install.packages("gapminder")
install.packages("ggridges")
install.packages("ggrepel")
install.packages("viridis")
install.packages("RColorBrewer")
install.packages("ggsci")
install.packages("plotly")
install.packages("gganimate")
install.packages("gifski")
```


Load R packages
```{r warning = FALSE, message = FALSE}
library(tidyverse) # a collection of data manipulation and visualization packages
library(gapminder) # package with a gapminder dataset
library(ggridges) # extension of ggplot2 that makes ridgeline plots
library(ggrepel) # extension of ggplot2 that makes non-overlapping labels
library(viridis) # color palette package
library(RColorBrewer) # color palette package
library(ggsci) # color palette package
library(plotly) # package for interactive data visualizations
library(gganimate) # package for producing gifs, extension of ggplot2
library(gifski) # package for converting video frames to GIF animations
```


# Preparing the data

Import data from the built-in "gapminder" dataset
```{r}
countries <- gapminder # create a data frame
head(countries) #see the first six rows of the dataframe
```


Check the data structure
```{r}
str(countries)
```


# Data manipulation with `dplyr`

## Filter function

Filter data of countries from 1997 and with Oceania continent
```{r}
filter(countries, year==1997 & continent=="Oceania") # Use "&" for AND condition
```


Filter observations from year 2002 or year 2007
```{r}
filter(countries, year==2002 | year==2007) # Use "|" for OR condition
```


## Summarise function

Available sub-functions:

* Center: mean(), median()
* Spread: sd(), IQR(), mad()
* Range: min(), max(), quantile()
* Position: first(), last(), nth(),
* Count: n(), n_distinct()
* Logical: any(), all()


Find average value of the lifeExp variable
```{r}
summarise(countries, mean_lifeExp=mean(lifeExp))
```


## Piping with multiple functions

What if you want to perform multiple functions in R? Use Pipe operator (%>%) in the dplyr package. It allows you to perform multiple functions without using nested parentheses.


This is how piping looks:
```{r}
# DataFrame%>%
    #function to execute first %>%
        #function to execute second %>%
            #function to execute third
```


## Group_by function

Group_by function allows you to do operations by groups. E.g. find maximum life expectancy by continent
```{r}
countries %>%
  group_by(continent) %>%
  summarise(max_lifeExp=max(lifeExp))
```


# Data visualization with `ggplot2`

## Barchart

Shows the distribution of a categorical variable
```{r}
countries %>%
  ggplot(aes(x=continent))+
    geom_bar()
```


Add unique colour to each continent using fill function
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar()
```


Add title, subtitle, x and y labels, and caption
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar()+
    labs(title="Distribution of observations by continents", 
         subtitle = "Measured as counts", 
         x="Continents",
         y="Number of observations",
         caption = "Source: Gapminder data")
```


Make a relative barchart by modifying y-values to measure the percentage of observations
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar(aes(y=stat(count)/sum(count)))+
    labs(title="Data distribution by continents", 
         x="Continents",
         y="Percent of observations")
```


Remove legend using `guides` command
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar(aes(y=stat(count)/sum(count)))+
    labs(title="Data distribution by continents", 
         x="Continents",
         y="Percent of observations")+
    guides(fill = "none")
```


## Histograms

Visualizes the distribution of one quantitative variable
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram()+
    labs(title="Histogram of life expenctancy ", 
         x="Life expectancy",
         y="Number of observations")
```


Use light-blue fill color with a black border
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram(color="black", fill="lightblue")+
    labs(title="Histogram of life expenctancy ", 
         x="Life expectancy",
         y="Number of observations")
```


Change bin-width from 2 to 4
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram(binwidth=4, color="black", fill="lightblue")+
    labs(title="Histogram of life expenctancy ", 
         x="Life expectancy",
         y="Number of observations")
```


Make a relative histogram by adding `y=..density..` (remember to change titles)
```{r}
countries %>%
  ggplot(aes(x=lifeExp,y=..density..))+
    geom_histogram(binwidth=4, color="black", fill="lightblue")+
    labs(title="Relative histogram of life expenctancy ", 
         x="Life expectancy",
         y="Percent of observations")
```


## Density plots


Visualizes the distribution of a quantitative variable
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_density()+
    labs(title="Density plot of life expenctancy", 
         x="Life expectancy")
```


Use pink fill color with a thicker border (size = 1)
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_density(size=1, fill="pink")+
    labs(title="Density plot of life expenctancy ", 
         x="Life expectancy")
```


Contrasting lifeExp distributions of continents using density plots
```{r}
countries %>%
  ggplot(aes(x=lifeExp, fill=continent))+
    geom_density()+
    labs(title="Density plot of life expenctancy by continent ", 
         x="Life expectancy")
```


Change alpha to make it more transparent, alpha measures color saturation (0 to 1)
```{r}
countries %>%
  ggplot(aes(x=lifeExp, fill=continent))+
    geom_density(alpha=0.3)+
    labs(title="Density plot of life expenctancy by continent ", 
         x="Life expectancy")
```


Or make a ridgeline plot using`geom_density_ridges` function (from `ggridges` package)
```{r}
countries %>%
  ggplot(aes(x=lifeExp, y=continent, fill=continent))+
    geom_density_ridges(alpha=0.7)+
    labs(title="Distribution of life expenctancy by continent ", 
         x="Life expectancy")
```


## Boxplots

Boxplots are used to compare distributions of one quantitative variable across multiple categories (a visualization alternative to density plots)


Make a boxplot of life expectancy
```{r}
countries %>%
  ggplot(aes(y=lifeExp))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy",
         y="Life expectancy")
```


Comparing distribution of lifeExp variable by continent
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")
```


Add unique colour to each continent
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_boxplot()+ 
    labs(title="Boxplots of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")
```


Reorder continents from smallest to highest life expectancy
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")
```


Change the name of your legend title to "Continent"
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",
         x="Continent",
         y="Life expectancy",
         fill="Continent")
```


Add average life expectancy values to each boxplot
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",
         x="Continent",
         y="Life expectancy",
         fill="Continent")+
    stat_summary(fun=mean, geom="point", shape=5, size=4)
```


## Violin charts

Make a violin chart of life expectancy grouped by continent
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent", 
         y="Life expectancy")
```

Add observations to violin plots using `geom_jitter`
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    geom_jitter(alpha=0.2, width=0.3)+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent", 
         y="Life expectancy")
```


## Small multiples

### `facet_wrap` function

Usinf `facet_wrap` function to break down each violin chart by year
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent", 
         y="Life expectancy")+
    facet_wrap(~year)
```


### `facet_grid` function

Stacking charts into columns with `facet_grid`
```{r}
countries %>%
  filter(year==1967 | year==2007) %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent", 
         y="Life expectancy")+
    facet_grid(.~year)
```


Stacking charts into rows with `facet_grid`
```{r}
countries %>%
  filter(year==1967 | year==2007) %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent", 
         y="Life expectancy")+
    facet_grid(year~.)
```


## Scatterplots


Scatterplots show the relationship between two quantitative variables
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point()+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="GDP per capita", 
         y="Life expectancy")
```


Add color to represent different continents
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent))+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="GDP per capita", 
         y="Life expectancy")
```


Change the size of data points to measure the size of population
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent, size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="GDP per capita", 
         y="Life expectancy")
```


### Transforming x and y values

Transform the gdpPercap variable using a `scale_x_log10()`
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_x_log10()
```


Specify ranges for x and y values using `xlim` and `ylim` functions
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_x_log10()+
    xlim(1000,10000)+
    ylim(50,70)
```


### Best fit lines

Add a best fit line to the scatterplot
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(method="lm", se=FALSE)
```


Add a best fit line for each continent and reduce saturation for points
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    labs(title="Scatterplot of GDP per capita and life expectancy", 
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(aes(color=continent),method="lm", se=FALSE)
```


### Exercise #1a

Add `facet_wrap`to create a small-multiple for each continent
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(color="black",method="lm",se=FALSE)

```

### Exercise #1b

Add `facet_grid`to create a small-multiple to compare continent between 1967 and 2007

  -  Use `facet_grid`and make years rows and continents columns
  
```{r}
countries %>%
  filter(year==1967|year==2007) %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(color=continent,size=pop), alpha=0.3)+
      labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
           x="Log 10 of GDP per capita", 
           y="Life expectancy")+
      scale_x_log10()+
      geom_smooth(color="black",method="lm",se=FALSE)

```

## Labels

Use `geom_text` function to add labels your observations
```{r}
countries %>%
  filter(year==2007, continent=="Europe") %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(size=pop), color="blue", alpha=0.3)+
      geom_text(aes(label=country), size = 3)+
      labs(title="GDP per capita and life expectancy in Europe in 2007",
           x="GDP per capita", 
           y="Life expectancy")
```

Use `geom_text_repel` function (from `ggrepel` package) to make sure labels do not overlap
```{r}
countries %>%
  filter(year==2007, continent=="Europe") %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(size=pop), color="blue", alpha=0.3)+
      geom_text_repel(aes(label=country), size = 3)+
      labs(title="GDP per capita and life expectancy in Europe in 2007",
           x="GDP per capita", 
           y="Life expectancy")
```

## Themes

  - `theme_gray()` - the signature ggplot2 theme with a grey background and white gridlines, designed to put the data forward yet make comparisons easy.
  - `theme_bw()` - the classic dark-on-light ggplot2 theme. May work better for presentations displayed with a projector.
  - `theme_linedraw()` - a theme with only black lines of various widths on white backgrounds, reminiscent of a line drawing. Serves a purpose similar to theme_bw. Note that this theme has some very thin lines (<< 1 pt) which some journals may refuse.
  - `theme_light()` - a  theme similar to theme_linedraw but with light grey lines and axes, to direct more attention towards the data.
  - `theme_dark()` - the dark cousin of theme_light, with similar line sizes but a dark background. Useful to make thin coloured lines pop out.
  - `theme_minimal()` - a minimalistic theme with no background annotations.
  - `theme_classic()` - a classic-looking theme, with x and y axis lines and no gridlines.
  - `theme()` - create your own theme by modifying theme components, see https://ggplot2.tidyverse.org/reference/theme.html


### Exercise #2

Play around with a few themes - contrast and compare, let us know your favorite ones in the comments
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    geom_smooth(aes(color=continent), method="lm",se=FALSE)+
    scale_x_log10()+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", 
         y="Life expectancy")

```


### Modify graphs with`theme()`
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    geom_smooth(aes(color=continent), method="lm",se=FALSE)+
    scale_x_log10()+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    theme_light()+
    theme( # change font size, font, and color for x-axis title
      axis.title.x=element_text(size=20, family="serif", face="italic", color="red"))
```



## Colors


### Set ggplot color manually

Use `scale_fill_manual()` for box plot, bar plot, violin plot, dot plot, etc
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    scale_fill_manual(values = c("#FFDB6D","#D16103","#52854C","#4E84C4","#293352"))
```


Use `scale_color_manual()` for lines and points
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    geom_smooth(aes(color=continent), method="lm",se=FALSE)+
    scale_x_log10()+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", 
         y="Life expectancy")+
    scale_color_manual(values = c("#FFDB6D","#D16103","#52854C","#4E84C4","#293352"))
```


### Viridis color palettes

The viridis R package provides color palettes to make beautiful plots that are: printer-friendly, perceptually uniform and easy to read by those with colorblindness.

  - Use `scale_fill_viridis()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_viridis()` for lines and points
  
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    scale_fill_viridis(discrete = TRUE)
```


### Rcolorbrewer palettes

RColorBrewer is an R package that contains a ready-to-use color palettes for creating beautiful graphics:

  - Sequential palettes (first list of colors), which are suited to ordered data that progress from low to high (gradient)
  - Qualitative palettes (second list of colors), which are best suited to represent nominal or categorical data
  - Diverging palettes (third list of colors), which put equal emphasis on mid-range critical values and extremes at both ends of the data range


Display available palettes
```{r}
display.brewer.all()
```


Display only colorblind-friendly brewer palettes
```{r}
display.brewer.all(colorblindFriendly = TRUE)
```


Which type of color palette is appropriate to visualize continents?

  - Use `scale_fill_brewer()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_brewer()` for lines and points
  
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    scale_fill_brewer(palette = "Set2")
```


### Grey color scales

  - Use `scale_fill_grey()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_grey()` for lines and points
  
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    scale_fill_grey(start = 1, end = 0.2) # ranges from 0 (black) to 1 (white)
```


### Scientific journal color palettes

The R package ggsci contains a collection of high-quality color palettes inspired by colors used in scientific journals, data visualization libraries, and more.

- scale_color_npg() and scale_fill_npg(): Nature Publishing Group color palettes
- scale_color_aaas() and scale_fill_aaas(): American Association for the Advancement of Science color palettes
- scale_color_lancet() and scale_fill_lancet(): Lancet journal color palettes
- scale_color_jco() and scale_fill_jco(): Journal of Clinical Oncology color palettes
- scale_color_tron() and scale_fill_tron(): This palette is inspired by the colors used in Tron Legacy. It is suitable for displaying data when using a dark theme.


### Exercise #3

Play around with different palettes - contrast and compare, let us know your favorite ones in the comments

Remember to use `fill` functions for box plot, bar plot, violin plot, dot plot
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")

```


## Annotations

Use `geom_hline` to add life expectancy in Canada as a reference line to your graph (replace for `geom_vline` for vertical lines):
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    geom_hline(color="red", yintercept=82.51)
```

Use `annotate` to add text and point to the line:
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",
         x="Continent",
         y="Life expectancy")+
    geom_hline(color="red", yintercept=82.51)+
    annotate("text", x = "Americas", y = 85.51, label = "Canada 2020", color="red")+
    annotate("point",x = "Americas", y = 82.51, color="black")
```


## Piping in `ggplot2`

Visualize changes in annual average life expectancy by continent with points representing the size of populations
```{r}
lifeExp_continent <- countries %>%
    group_by(continent, year) %>%
      summarise(lifeExp=mean(lifeExp), pop=sum(pop)) %>%
        ggplot(aes(x=year, y=lifeExp, color=continent))+
          geom_point(aes(size=pop))+
          labs(title="Annual average life expectancy by continent", 
               y="Life expectancy")+
          geom_line()

lifeExp_continent
```


## Save produced graphs

Use `ggsave` function to save the plot in your working directory
```{r}
ggsave("lifeExp_continent.png", plot = lifeExp_continent, device="png", width = 8, height = 5)
```


# Interactive visualizations

## `ggplotly`

Make an interactive plotly scatterplot of life expectancy and GDP per capita in 2007
```{r}
plot1 <- countries %>%
  filter(year==2007) %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(size=pop, color=continent, label=country))+
      labs(title="Life expectancy and GDP per capita, 2007",
           x="Log 10 of GDP per capita", 
           y="Life expectancy")+
      scale_x_log10()+
      geom_smooth(aes(color=continent), method="lm", se=FALSE)
ggplotly(plot1)
```


Add year as a frame so that we can see changes over time
```{r}
plot2 <- countries %>%
    ggplot(aes(x=gdpPercap, y=lifeExp, frame=year))+
      geom_point(aes(color=continent, size=pop,label=country))+
      geom_smooth(aes(color=continent), method="lm",se=FALSE)+
      scale_x_log10()+
      labs(title="Life exp. and GDP per capita, 1952-2007",
           x="Log 10 of GDP per capita", 
           y="Life expectancy")
ggplotly(plot2)
```

Note: You can make any static `ggplot2` graph interactive with `plotly`


## `gganimate`

Create an animation using `transition_time` function + show historical traces by adding `shadow_wake` function
```{r}
gif_countries <- countries %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(color=continent, size=pop,label=country))+
      scale_x_log10()+
      transition_time(year)+
      shadow_wake(wake_length = 0.1)+
      labs(title="Life expectancy and GDP per capita in {frame_time}",
           x="Log 10 of GDP per capita", 
           y="Life expectancy")

gif_countries
```


Save animation using `anim_save` function
```{r}
anim_save("countries.gif", animation = gif_countries, path = NULL)
```
