---
title: "Visualization with R workshop"
author: "Jeremy Buhler and Albina Gibadullina"
date: "February 28, 2021"
output: html_document
---

# Visualization in R

## Installing and loading packages

Install R packages
```{r}
install.packages("dplyr")
install.packages("gglot2")
install.packages("gapminder")
install.packages("forcats")
install.packages("plotly")
install.packages("viridis")
install.packages("RColorBrewer")
install.packages("ggsci")
```


Load R packages
```{r}
library(dplyr) # data manipulation package
library(ggplot2) # data visualization package
library(gapminder) # package with a gapminder dataset
library(forcats) # package for working with categorical data
library(plotly) # package for interactive data visualizations
library(viridis) # color palette package
library(RColorBrewer) # color palette package
library(ggsci) # color palette package
```


## Preparing the data

Import data from the built-in "gapminder" dataset
```{r}
countries <- gapminder # create a data frame
head(countries) #see the first six rows of the dataframe
```


Check the data structure
```{r}
str(countries)
```


## Review of data manipulation with `dplyr`

### Filter function

Filter data of countries from 1997 and with Oceania continent
```{r}
filter(countries, year==1997, continent=="Oceania") # Use "," for AND condition
```


Filter observations from year 2002 or year 2007
```{r}
filter(countries, year==2002 | year==2007) # Use "|" for OR condition
```


### Summarise function

Available sub-functions:

* Center: mean(), median()
* Spread: sd(), IQR(), mad()
* Range: min(), max(), quantile()
* Position: first(), last(), nth(),
* Count: n(), n_distinct()
* Logical: any(), all()


Find average value of the lifeExp variable
```{r}
summarise(countries, mean_lifeExp=mean(lifeExp))
```


### Piping with multiple functions

What if you want to perform multiple functions in R? Use Pipe operator (%>%) in the dplyr package. It allows you to perform multiple functions without using nested parentheses.


This is how piping looks:
```{r}
# DataFrame%>%
    #function to execute first %>%
        #function to execute second %>%
            #function to execute third
```


### Group_by function

Group_by function allows you to do operations by groups. E.g. find maximum life expectancy by continent
```{r}
countries %>%
  group_by(continent) %>%
  summarise(max_lifeExp=max(lifeExp))
```


# Data visualization with `ggplot2`

## Barchart

Shows the distribution of a categorical variable
```{r}
countries %>%
  ggplot(aes(x=continent))+
    geom_bar()
```


Add unique colour to each continent using fill function
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar()
```


Add title and x and y labels
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar()+
    labs(title="Counts of observations by continents", x="Continents",
         y="Number of observations")
```


Make a relative barchart by modifying y-values to measure the percentage of observations
```{r}
countries %>%
  ggplot(aes(x=continent, fill=continent))+
    geom_bar(aes(y=stat(count)/sum(count)))+
    labs(title="Data distribution by continents", x="Continents",
         y="Percent of observations")
```


## Histograms

Visualizes the distribution of one quantitative variable
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram()+
    labs(title="Histogram of life expenctancy ", x="Life expectancy",
       y="Number of observations")
```


Use light-blue fill color with a black border
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram(color="black", fill="lightblue")+
    labs(title="Histogram of life expenctancy ", x="Life expectancy",
       y="Number of observations")
```


Change bin-width from 2 to 4
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_histogram(binwidth=4, color="black", fill="lightblue")+
    labs(title="Histogram of life expenctancy ", x="Life expectancy",
       y="Number of observations")
```


## Alternative to histograms - density plots


Visualizes the distribution of a quantitative variable
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_density()+
    labs(title="Density plot of life expenctancy ", x="Life expectancy")
```


Use pink fill color with a thicker border (size = 1)
```{r}
countries %>%
  ggplot(aes(x=lifeExp))+
    geom_density(size=1, fill="pink")+
    labs(title="Density plot of life expenctancy ", x="Life expectancy")
```


Contrasting lifeExp distributions of continents using density plots
```{r}
countries %>%
  ggplot(aes(x=lifeExp, fill=continent))+
    geom_density()+
    labs(title="Density plot of life expenctancy by continent ", x="Life expectancy")
```


Change alpha to make it more transparent, alpha measures color saturation (0 to 1)
```{r}
countries %>%
  ggplot(aes(x=lifeExp, fill=continent))+
    geom_density(alpha=0.3)+
    labs(title="Density plot of life expenctancy by continent ", x="Life expectancy")
```


## Boxplots

Boxplots are used to compare distributions of one quantitative variable across multiple categories (a visualization alternative to density plots)


Make a boxplot of life expectancy
```{r}
countries %>%
  ggplot(aes(y=lifeExp))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy",
         y="Life expectancy")
```


Comparing distribution of lifeExp variable by continent
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",x="Continent",
         y="Life expectancy")
```


Add unique colour to each continent
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_boxplot()+ 
    labs(title="Boxplots of life expenctancy by continent",x="Continent",
         y="Life expectancy")
```


Reorder continents from smallest to highest life expectancy
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",x="Continent",
         y="Life expectancy")
```


Change the name of your legend title to "Continent"
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",x="Continent",
         y="Life expectancy",fill="Continent")
```


Add average life expectancy values to each boxplot
```{r}
countries %>%
  ggplot(aes(x=reorder(continent,lifeExp), y=lifeExp, fill=continent))+
    geom_boxplot()+
    labs(title="Boxplots of life expenctancy by continent",x="Continent",
         y="Life expectancy",fill="Continent")+
    stat_summary(fun=mean, geom="point", shape=5, size=4)
```


## Alternatives to boxplots - Violin charts


Make a violin chart of life expectancy grouped by continent

```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent", y="Life expectancy")
```


## Small multiples

### `facet_wrap` function

Usinf `facet_wrap` function to break down each violin chart by year
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent", y="Life expectancy")+
    facet_wrap(~year)
```


### `facet_grid` function

Stacking charts into columns with `facet_grid`
```{r}
countries %>%
  filter(year==1967 | year==2007) %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent", y="Life expectancy")+
    facet_grid(.~year)
```


Stacking charts into rows with `facet_grid`
```{r}
countries %>%
  filter(year==1967 | year==2007) %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent", y="Life expectancy")+
    facet_grid(year~.)
```


## Scatterplots


Scatterplots show the relationship between two quantitative variables
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point()+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="GDP per capita", y="Life expectancy")
```


Add color to represent different continents
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="GDP per capita", y="Life expectancy")
```


Change the size of data points to measure the size of population
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent, size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="GDP per capita", y="Life expectancy")
```


### Transforming x and y values

Transform the gdpPercap variable using a `scale_x_log10()`
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_x_log10()
```


Specify ranges for x and y values using `xlim` and `ylim` functions
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_x_log10()+
    xlim(1000,10000)+
    ylim(50,70)
```


### Best fit lines

Add a best fit line to the scatterplot
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(method="lm", se=FALSE)
```


Add a best fit line for each continent
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop))+
    labs(title="Scatterplot of GDP per capita and life expectancy", x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(aes(color=continent),method="lm", se=FALSE)
```



### Exercise #1a:

Add `facet_wrap`to create a small-multiple for each continent
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_x_log10()+
    geom_smooth(color="black",method="lm",se=FALSE)

```

### Exercise #1b:

Add `facet_grid`to create a small-multiple to compare continent between 1967 and 2007

  -  Use `facet_grid`and make years rows and continents columns

```{r}
countries %>%
  filter(year==1967|year==2007) %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(color=continent,size=pop), alpha=0.3)+
      labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
           x="Log 10 of GDP per capita", y="Life expectancy")+
      scale_x_log10()+
      geom_smooth(color="black",method="lm",se=FALSE)

```



## THEMES

  - `theme_gray()` - the signature ggplot2 theme with a grey background and white gridlines, designed to put the data forward yet make comparisons easy.
  - `theme_bw()` - the classic dark-on-light ggplot2 theme. May work better for presentations displayed with a projector.
  - `theme_linedraw()` - a theme with only black lines of various widths on white backgrounds, reminiscent of a line drawing. Serves a purpose similar to theme_bw. Note that this theme has some very thin lines (<< 1 pt) which some journals may refuse.
  - `theme_light()` - a  theme similar to theme_linedraw but with light grey lines and axes, to direct more attention towards the data.
  - `theme_dark()` - the dark cousin of theme_light, with similar line sizes but a dark background. Useful to make thin coloured lines pop out.
  - `theme_minimal()` - a minimalistic theme with no background annotations.
  - `theme_classic()` - a classic-looking theme, with x and y axis lines and no gridlines.
  - `theme()` - create your own theme by modifying theme components, see https://ggplot2.tidyverse.org/reference/theme.html


### Exercise #2

Play around with a few themes - contrast and compare, let us know your favorite ones in the comments
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    geom_smooth(aes(color=continent), method="lm",se=FALSE)+
    scale_x_log10()+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", y="Life expectancy")

```



## COLOURS


### Set ggplot color manually

Use `scale_fill_manual()` for box plot, bar plot, violin plot, dot plot, etc
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent",y="Life expectancy")+
    scale_fill_manual(values = c("#FFDB6D","#D16103","#52854C","#4E84C4","#293352"))
```


Use `scale_color_manual()` for lines and points
```{r}
countries %>%
  ggplot(aes(x=gdpPercap, y=lifeExp))+
    geom_point(aes(color=continent,size=pop), alpha=0.3)+
    geom_smooth(aes(color=continent), method="lm",se=FALSE)+
    scale_x_log10()+
    labs(title="Best-fit lines between gdpPercap(log10) and lifeExp for each continent",
         x="Log 10 of GDP per capita", y="Life expectancy")+
    scale_color_manual(values = c("#FFDB6D","#D16103","#52854C","#4E84C4","#293352"))
```


### Viridis color palettes

The viridis R package provides color palettes to make beautiful plots that are: printer-friendly, perceptually uniform and easy to read by those with colorblindness.

  - Use `scale_fill_viridis()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_viridis()` for lines and points
  
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent",y="Life expectancy")+
    scale_fill_viridis(discrete = TRUE)
```


### Rcolorbrewer palettes

RColorBrewer is an R package that contains a ready-to-use color palettes for creating beautiful graphics:

  - Sequential palettes (first list of colors), which are suited to ordered data that progress from low to high (gradient)
  - Qualitative palettes (second list of colors), which are best suited to represent nominal or categorical data
  - Diverging palettes (third list of colors), which put equal emphasis on mid-range critical values and extremes at both ends of the data range


Display available palettes
```{r}
display.brewer.all()
```


Display only colorblind-friendly brewer palettes
```{r}
display.brewer.all(colorblindFriendly = TRUE)
```


Which type of color palette is appropriate to visualize continents?

  - Use `scale_fill_brewer()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_brewer()` for lines and points
  
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent",y="Life expectancy")+
    scale_fill_brewer(palette = "Set2")
```


### Grey color scales

  - Use `scale_fill_grey()` for box plot, bar plot, violin plot, dot plot
  - Use `scale_color_grey()` for lines and points
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent",y="Life expectancy")+
    scale_fill_grey(start = 1, end = 0.2) # ranges from 0 (black) to 1 (white)
```


### Scientific journal color palettes

The R package ggsci contains a collection of high-quality color palettes inspired by colors used in scientific journals, data visualization libraries, and more.

- scale_color_npg() and scale_fill_npg(): Nature Publishing Group color palettes
- scale_color_aaas() and scale_fill_aaas(): American Association for the Advancement of Science color palettes
- scale_color_lancet() and scale_fill_lancet(): Lancet journal color palettes
- scale_color_jco() and scale_fill_jco(): Journal of Clinical Oncology color palettes
- scale_color_tron() and scale_fill_tron(): This palette is inspired by the colors used in Tron Legacy. It is suitable for displaying data when using a dark theme.


#### Exercise #3:

Play around with different palettes - contrast and compare, let us know your favorite ones in the comments

Remember to use `fill` functions for box plot, bar plot, violin plot, dot plot
```{r}
countries %>%
  ggplot(aes(x=continent, y=lifeExp, fill=continent))+
    geom_violin()+
    labs(title="Distribution of life expenctancy by continent",x="Continent",y="Life expectancy")

```


## Piping in ggplot

Visualize changes in annual average life expectancy by continent with points representing the size of populations
```{r}
countries %>%
    group_by(continent, year) %>%
      summarise(lifeExp=mean(lifeExp), pop=sum(pop)) %>%
        ggplot(aes(x=year, y=lifeExp, color=continent))+
          geom_point(aes(size=pop))+
          labs(title="Annual average life expectancy by continent", y="Life expectancy")+
          geom_line()
```


### Save produced graphs

If you want to save the produced graph, you need to assign it to an object with a new name
```{r}
lifeExp_continent <- countries %>%
    group_by(continent, year) %>%
      summarise(lifeExp=mean(lifeExp), pop=sum(pop)) %>%
        ggplot(aes(x=year, y=lifeExp, color=continent))+
          geom_point(aes(size=pop))+
          labs(title="Annual average life expectancy by continent", y="Life expectancy")+
          geom_line()

lifeExp_continent
```


Use `ggsave` function to save the plot in your working directory
```{r}
ggsave("lifeExp_continent.png", plot = lifeExp_continent, device="png", width = 8, height = 5)
```



## PLOTLY

Make an interactive plotly scatterplot of life expectancy and GDP per capita in 2007
```{r}
plot1 <- countries %>%
  filter(year==2007) %>%
    ggplot(aes(x=gdpPercap, y=lifeExp))+
      geom_point(aes(size=pop, color=continent, label=country))+
      labs(title="Life expectancy and GDP per capita, 2007",x="Log 10 of GDP per capita", y="Life expectancy")+
      scale_x_log10()+
      geom_smooth(aes(color=continent), method="lm",se=FALSE)

ggplotly(plot1)
```


Add year as a frame so that we can see changes over time
```{r}
plot2 <- countries %>%
    ggplot(aes(x=gdpPercap, y=lifeExp, frame=year))+
      geom_point(aes(color=continent, size=pop,label=country))+
      geom_smooth(aes(color=continent), method="lm",se=FALSE)+
      scale_x_log10()+
      labs(title="Life exp. and GDP per capita, 1952-2007",
           x="Log 10 of GDP per capita", y="Life expectancy")

ggplotly(plot2)
```


You can make any static `ggplot2` graph interactive with `plotly`
```{r}
violin <- countries %>%
    ggplot(aes(x=continent, y=lifeExp, fill=continent))+
      geom_violin()+
      geom_jitter(aes(label=country),width = 0.2, size=0.2, alpha=0.4)+
      labs(title="Distribution of life exp. by continent and year",
           x="Continent",y="Life expectancy")+
      facet_wrap(~year)

ggplotly(violin)
```


## Save the dataframe as a CSV file

Save the dataframe as a csv file
```{r}
write.csv(countries, file="countries.csv")
```
