---
title: "Advanced Skills Workshop"
output: html_document
date: "2022-11-21"
---

# Workshop overview: 

- For loops
- Nested for loops 
- Break, 5-10 mins 
- Conditional (if/else) statements
- Custom functions 
- Additional practice problem

# 1 GETTING STARTED 

```{r}
#install.packages('openintro')
#install.packages('dplyr')
#install.packages('gapminder')
#install.packages('ggplot2')
#install.packages('viridis)

library(openintro)
library(dplyr)
library(gapminder)
library(ggplot2)
library(viridis)

```

We will be using three datasets in this workshop: 

(1) 'countries' from the gapminder package - Macroeconomic (e.g. GDP per capita) and demographic (e.g. population) data for every country in the world, by year 


```{r}
countries <- gapminder # create a data frame
head(countries) #see the first six rows of the dataframe
str(countries) #look at the structure of the dataframe
names(countries) #variables in the dataset
```

(2) 'storms' from the dplyr package - Location and weather data for hurricanes in the US

```{r}
head(storms)
```

(3) 'midterms' from the openintro package - midterm election results by year, party, and unemployment rate 

```{r}
midterms <- midterms_house
head(midterms)
?midterms_house #look at dataset documentation 
```

# 2 FOR LOOPS

## 2.1 Basic structure 

for (i in vector/list/matrix/dataframe){
 some operation with i 
}

one step in a loop = one 'iteration' 

numeric vector example: 
```{r}
#a simple loop: 

numbers <- 1:10 #creating a vector of numbers 1 through 10 

for (i in numbers){
  print(i)
}

for (i in numbers){
  print(i)
  Sys.sleep(0.5) #this staggers each output by 0.5 seconds
}

#in this illustrative example, you need to include print, or the loop will run without producing output. 
```

numeric vector example: 
```{r}
other_numbers <- c(4,7,5,2,2,3,1)

for (i in other_numbers){
  print(i) #i is looped through in the given order 
}
```

character vector example: 
```{r}
letters #displaying built-in alphabet vector 

for (i in letters){
  print(i)
}
```
```{r}
matrix <- matrix(nrow = 3, ncol = 5, c(1:15))
matrix

for (i in matrix){
  print(i)
}
```

dataframe example: 
```{r}
###looping through the unique political parties in our midterms dataset
for (i in unique(midterms$party)){
  print(i)
}

#note i is an object and never in quotes. i is commonly-used 
#but arbitrary. You can make it more meaningful to you if thats easier, e.g.: 

for (Parties in unique(midterms$party)){
  print(Parties)
}
```

Applying an operation by looping through the parties in the midterms dataset
```{r}
###finding the maximum value of house_change for each party
#if we wanted to do this for ONE party, e.g. Republican 

max(midterms$house_change[midterms$party == 'Republican'])

#uses square brackets to subset, will be using this throughout the workshop
```

individual components we need to build our for loop: 

1. unique(midterms$party)
2. max(midterms$house_change[midterms$party == Party])
3. for loop struture: 

for (i in vector){
 operation 
}

```{r}
#building this into a for loop, for all parties
for (i in unique(midterms$party)){
  x <- max(midterms$house_change[midterms$party == i])
  print(x)
}

#can save as an object or embed max within print , i.e 
#print(max(countries$gdpPercap[countries$country == i]))
```

### 2.1.1 PRACTICE TOGETHER 

Another example, with a different dataset: 
finding the average latitude for each storm in the dataset

individual components we need to build our for loop: 

1. unique(storms$name) 
2. mean(storms$lat[storms$name == Storm name])
3. for loop struture: 

for (i in vector){
 operation 
}

```{r}
#storms dataset, from dplyr
head(storms)

#finding the average latitude for each storm in the dataset 

```

PRACTICE: 

Using the 'countries' dataset, use a for loop to find the maximum value of GDP per capita (gdpPercap) for each country

```{r}
#answer: 

```

## 2.2 Saving the output of your for loop:

lets create a new dataset called 'country_sum, of summary variables by country, with country as one column and max gdp per capita as another 

```{r}
#creating empty vectors for the values we want to store
country <- 0
maxgdp <- 0

#to store values in a vector, we need to create a count, to add values to a vector as we move through each iteration of the loop 
index <- 0 

#vector[x] points to position x in the vector

for (i in unique(countries$country)){
  maxgdp[index] <- max(countries$gdpPercap[countries$country == i])
    #create vector of max gdpPercap values by country 
  country[index] <- i #create vectors of countries
  index <- index + 1 #add one to the index to avoid replacement
  print(index) #print index at each iteration (not required)
}

#look at our new vectors
head(maxgdp)
head(country)

#combine vectors into a dataframe
country_sum <- data.frame(country, maxgdp)
head(country_sum)

#there are other ways of doing this, by creating a dataframe and using that directly in the for-loop, but I think this format is more beginner-friendly 
```

### 2.2.1 PRACTICE: 

modify the code below to create a new dataframe with two columns: Party, and max midterms house change (x in the example code). Name your columns however you'd like. 

```{r}

#original code:
for (i in unique(midterms$party)){
  x <- max(midterms$house_change[midterms$party == i])
  print(x)
}

#answer: 

```

More complicated operations in looping 
```{r}
#add year 

year <- 0 
country <- 0
maxgdp <- 0

index <- 0 

for (i in unique(countries$country)){
  maxgdp[index] <- max(countries$gdpPercap[countries$country == i])
  country[index] <- i #same as previous example up to here 
  year[index] <- countries$year[countries$gdpPercap == maxgdp[index]]
  index <- index + 1 #add one to the index to avoid replacement
}

country_sum <- data.frame(country, maxgdp, year)
head(country_sum)

```

## 2.3 NESTED FOR LOOPS 

nice analogies for nested looping: 

  - like going through words in a book, by pages 
  - like going through values in rows of a matrix, by columns 
  
```{r}

#conceptual example: 

page1 <- c("I", "wrote", "this", "book")
page2 <- c("I", "think", "its", "pretty", "good")
page3 <- c("The", "end")

mybook <-list(page1, page2, page3)

for (i in 1:length(mybook)){
  print(i)
  for (j in mybook[[i]]){
    print(j)
  }
}
```

example with the storms dataset 
```{r}
head(storms)

#Lets create a new dataframe, where we summarize the total number of days of storms, by 'status' (type of storm), for each year of data

#creating empty vectors for our desired variables 

days <- 0 
year <- 0 
status <- 0 

index <-0 

for (i in unique(storms$year)){ #looping through years
  for (j in unique(storms$status)){ #looping through storm status
    days[index] <- nrow(storms[storms$year == i & storms$status == j,])/4 #counting the number of rows of data in each year, for each status to get to the number of days 
#have to divide by four, because there are four entries per day
    year[index] <- i 
    status[index] <- j
    index <- index + 1 
  }
}

storm_data <- data.frame(year, status, days)
head(storm_data)
```

working with your new dataset: 
```{r}
#can plot your data 

ggplot(storm_data, aes(x = year, y = days, col = status)) + 
  geom_point() + 
  geom_smooth(aes(col=status),method="lm", se = FALSE)+
  #change colours
  scale_colour_viridis(discrete = TRUE) + 
  #rotate axis text, modifying theme 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  theme_minimal()
```

```{r}
#or analyze your data 

model_list<-list() #creating empty list

#create a list of models 
for (i in unique(storm_data$status)){
  model_list[[i]] <- summary(lm(days ~ year, data = storm_data[storm_data$status == i,]))
}

model_list
```

```{r}
#or create individual objects for each model based on i 
for (i in unique(storm_data$status)){
  assign(paste0("model_", i), summary(lm(days ~ year, data = storm_data[storm_data$status == i,])))
}

model_hurricane

```

# BREAK 

# 3 IF/ELSE STATEMENTS  

## 3.1 Basic structure

  if (logical statement){
   operation true} 
   
can be expanded to an if/else statement: 
  
  if (logical statement){
   operation if true} 
   else{operation if false}
   
examples of logical statements: 

x > 5 
x == 4 
x == "cat"
x != 0

Simple example: 

```{r}

x = 6 #set x to 6 

if (x > 5){
  print("x is greater than 5")
}

x = 4 #set x to 4 

#add an else statement 
if (x > 5){
  print("x is greater than 5")
} else {
  print("x is less than 5")
}
```

if/else statement become very useful when embedded within a for loop: 

```{r}

#e.g., create a binary variable for Republican (1) and Democrat (0)

midterms$binary <- NA

for (i in 1:nrow(midterms)){
if(midterms$party[i] == 'Republican'){
  midterms$binary[i] = 1} else {
    midterms$binary[i] = 0
  }
}

head(midterms)
```

# 4 CUSTOM FUNCTIONS

## 4.1 Basic structure 

my_function <- function(argument1, argument2, ...) {
  some operation using arguments 
  return(desired output)
}

for example, building a 'multiplier' function:

```{r}

multiplier <- function(n1, n2){
  output <- n1*n2
  return(output)
}

#or 

multiplier <- function(n1, n2){
  return(n1*n2)
}

multiplier(2,5)

```

### 4.1.1 PRACTICE (5-10 mins): 

build a function (called 'my_mean') that calculates the mean of two numbers (n1 and n2), without using the mean function

```{r}
#answer:

```

if we wanted to expand this, to work on an entire vector: 

```{r}

vec <- c(1,3,4,7,4,3) #creating new vector 

my_mean <- function(vec){
  return((sum(vec))/length(vec))
}

my_mean(vec) #runnning our function on our vector 
mean(vec) #checking it worked 
```

more complicated example, building a standard deviation function: 

standard deviation =  sqrt(sum((x - mean)^2)/N)

```{r}

#to do this OUTSIDE a custom function 

numbers<-c(2,4,5,1,3,4)

x <- 0
index <- 1 

for (i in numbers){
  x[index] <- (i - mean(numbers))^2
  index <- index + 1 
}

stddev <- sqrt(sum(x)/length(numbers))
stddev
```

if we wanted to repeat our calculation for many different vectors, we could create a custom function with a for loop 

```{r}

standard.dev <- function(values){ #providing our function argument
x <- 0 #creating and empty vector 
index <- 1 #creating an index for storing values
for (i in values){
  x[index] <- (i - mean(values))^2
  index <- index + 1 
}
stddev <- sqrt(sum(x)/length(values)) #note output goes outside 
return(stddev) #the for loop 
}
  
standard.dev(values = midterms$unemp)
#or 
standard.dev(midterms$unemp)
```

### 4.1.2 EXTRA CHALLENGE

modify your mean function to calculate the mean of each vector in the list provided (note to grab one vector from a list, you would use [[ ]], eg list[[2]] would grab the second vector in a list)

```{r}
vec1 <- c(1,2)
vec2 <- c(4,5,6)
vec3 <- c(7,8,9,10)

vec_list <- list(vec1, vec2, vec3)

my_mean_list <- function(list1){
  output <- 0 
    for (i in 1:length(list1)){
    output[i] <- (sum(list1[[i]])/length(list1[[i]]))}
    return(output)
}

my_mean_list(vec_list)
```

