---
layout: default
title: Markdown file
nav_order: 5
---

---
title: "Intro to R and RStudio Workshop"
author: "Albina Gibadullina and Maria Laricheva"
date: "September 24, 2021"
output: 
  html_document:
    theme: yeti
    toc: true
    toc_float: true
---

# Intro to R and RStudio

## Installing and loading needed packages

Install R packages
```{r}
#install.packages("tidyverse")
#install.packages("gapminder")
```

Load R packages
```{r}
library(tidyverse) # a collection of R packages designed for data science (e.g.dplyr, ggplot2, tidyr, forcats)
library(gapminder) # a package with the gapminder dataset
```


## Manually inputting and/or modifying values

### Creating new variables

Assign values to variables
```{r}
a <- 4
a
b <- 5
b
```

Create a new variable from existing variables
```{r}
c <- 6+a
c
d <- a + b + c
d
```

See if two variables equal to one another
```{r}
c==d
```

See if two variables are not equal to one another
```{r}
c!=d
```


### Working with vectors

Create a vector
```{r}
vec <- c(0, 1, 2, 3, 4, 5)
vec
```


Select the first value of vec:
```{r}
vec[1]
```


And here is how you can modify it:
```{r}
vec[1] <- 1000
vec
```


You can replace multiple values at once as long as the number of new values equals the number of selected values:
```{r}
vec[c(1, 3, 5)] <- c(-1, -2, -3)
vec
```

Add +1 to column 4 to 6
```{r}
vec[4:6] <- vec[4:6] + 1
vec
```

You can also create values that do not yet exist in your object. R will expand the object to accommodate the new values:
```{r}
vec[7] <- 10
vec
```


### Working with data frames

Create a data frame with four variables
```{r}
df <- data.frame(c("Anna","Patrick","John"),c(165,170,168),c(18000,45000,57000),c("youth","senior","adult"))
names(df) <- c("Name","Height", "Income", "Age")
df
```


Extract all rows, all colums
```{r}
df[,]
```

Extract first row and all columns
```{r}
df[1,]
```

Extract first two rows and all columns
```{r}
df[1:2,]
```

Extract first and third row and all columns
```{r}
df[c(1,3),]
```

Extract first row and second and third column
```{r}
df[1,2:3]
```


### Exercise #1

Extract third column with all rows
```{r}

```

Extract element at 2nd row, 3rd column
```{r}

```

Substitute element at 2nd row, 3rd column with 50000
```{r}

```


You can also extract values using variable names by adding `$` and variable name after the dataset name

E.g. Extract the first two observations of the `Age` variable
```{r}
df$Age[1:2]
```


## Changing format of variables

Main data types:

* character (text/strings)
* factor (categorical variable)
* numeric (quantitative variable)
* integer (quantitative variable, only integers)
* logical (TRUE or FALSE variable)


Check current data format of the  `Age` variable
```{r}
class(df$Age)
```

Change the format of `Age` from charachter to factor (categorical)
```{r}
df$Age <- as.factor(df$Age)
```

Check whether `Age` has factor format
```{r}
is.factor(df$Age)
```

Check levels for `Age`
```{r}
levels(df$Age)
```

If you wanted to re-arrange the order of levels: "youth", "adult", "senior"
```{r}
df$Age <- fct_relevel(df$Age, c("youth", "adult", "senior"))
#Check levels for `gender`
levels(df$Age)
```

### Commands to change data type
* as.numeric (quantitative, continious format) 
* as.integer (quantitative, discrete format)
* as.charachter (text format)
* as.factor (categorical format)
* as.logical (logical format: TRUE/FALSE)

### Commands to check your data type
* is.numeric (quantitative, continious format) 
* is.integer (quantitative, discrete format)
* is.charachter (text format)
* is.factor (categorical format)
* is.logical (logical format: TRUE/FALSE)


## Importing data 

### Importing data from external sources

CSV data files
```{r}
#countries <- read.csv(file="countries.csv") # make sure the file is located in your working directory
```


SPSS, SAS, or Stata data file
```{r}
#install.packages("foreign")
#library(foreign)
#countries <- read.spss(file="countries.sav") # SPSS
#countries <- read.dta(file="countries.dta") # Stata
#countries <- read.ssd(file="countries.sas") # SAS
```



### Working with built-in datasets

Check the description of the "gapminder" dataset
```{r}
#?gapminder
```


Import data from the built-in "gapminder" dataset
```{r}
countries <- gapminder # create a data frame
head(countries) #see the first six rows of the dataframe 
```


## See the overview of data

See the entire dataset
```{r}
View(countries)
```


Check the data structure
```{r}
str(countries)
```

Count the number of observations (measured by the number of rows) and the number of variables (measured by the number of columns)
```{r}
nrow(countries) #To check the number of rows
ncol(countries) #To check the number of columns
```


Summarise the dataframe
```{r}
summary(countries)
```

Check if the dataframe is complete:

Count the number of observations for each available year
```{r}
table(countries$year)
```


Count the number of observations in each continent for each available year 
```{r}
table(countries$continent, countries$year)
```


## Data Manipulation with `dplyr`


### Select function

Select specific columns such as country, year, lifeExp
```{r}
select(countries, country, year, lifeExp)
```


Select all columns except for continent and gdpPercap
```{r}
select(countries, -continent, -gdpPercap) # use subtraction operator
```

### Arrange function

Arrange countries based on life expectancy (from smallest to largest)
```{r}
arrange(countries, lifeExp)
```


Arrange countries based on life expectancy (from smallest to largest), sorted first by year (from earliest year available to most recent)
```{r}
arrange(countries, year, lifeExp)
```


Arrange countries based on life expectancy (from largest to smallest)
```{r}
arrange(countries, desc(lifeExp))
```


Arrange countries based on life expectancy (from largest to smallest), sorted first by year (from earliest year available to most recent)
```{r}
arrange(countries, year, desc(lifeExp))
```


### Filter function

Filter data of countries from 1997 and with Oceania continent
```{r}
filter(countries, year==1997, continent=="Oceania")
```


Filter data of countries from 2007 and with gdpPercap greater than or equal to $40,000
```{r}
filter(countries, year==2007, gdpPercap>=40000)
```


Filter observations from year 2002 or year 2007
```{r}
filter(countries, year==2002 | year==2007) # Use vertical line for Or condition
```


### Mutate function

Create a new variable called `GDP` (a product of `gdpPercap` and `pop`)
```{r}
mutate(countries, GDP=gdpPercap*pop)
```


#### Paste function

Create a new variable called `country_year` using `paste` function
```{r}
mutate(countries, country_year=paste(country,year,sep = " "))
```


#### Replace and ifelse functions

Create a dummy variable called `europe`
```{r}
mutate(countries, europe=replace(continent=="Europe",1,0))
```

Create a variable called `decade`
```{r}
mutate(countries, decade=ifelse(year<1970,"Before 1970",
                         ifelse(year>=1970 & year <1990, "1970 to 1990","After 1990")))
```


### Rename function

Rename`pop` variable as `population`
```{r}
rename(countries, population = pop)
```


### Summarise function

Available sub-functions:

* Center: mean(), median()
* Spread: sd(), IQR(), mad()
* Range: min(), max(), quantile()
* Position: first(), last(), nth(),
* Count: n(), n_distinct()
* Logical: any(), all()


Find average value of the lifeExp variable
```{r}
summarise(countries, mean_lifeExp=mean(lifeExp))
```


Find range of the lifeExp variable
```{r}
summarise(countries, range_lifeExp=max(lifeExp)-min(lifeExp))
```


### Piping with Multiple Functions

What if you want to perform multiple functions in R? Use Pipe operator (%>%) in the dplyr package. It allows you to perform multiple functions without using nested parentheses. 


This is how piping looks like:
```{r}
# DataFrame%>%
    #function to execute first %>%
        #function to execute second %>%
            #function to execute third
```


#### Group_by function

Group_by function allows you to do operations by groups. E.g. find maximum life expectancy by continent
```{r}
countries %>% 
  group_by(continent) %>% 
  summarise(max_lifeExp=max(lifeExp))
```


#### Example of Piping

Finding average life expectancy in each continent in 2007

* Select country, continent, year, lifeExp
* Filter observations from 2007
* Split data by continents
* Summarise the mean life expectancy within each continent
* Round the average life expectancy to 2 decimal points


```{r}
countries %>%
  select(country, continent, year, lifeExp) %>% #select country, continent,  year, lifeExp
    filter(year==2007) %>% #filter observations from 2007
      group_by(continent) %>% # split data by continents
          summarise(mean_lifeExp=mean(lifeExp)) %>% #summarise the mean life expectancy within each continent
            mutate(mean_lifeExp=round(mean_lifeExp,2)) # Use mutate to modify existing variable, round mean weight to 2 decimal points
```


### Exercise #2

Finding the average of GDP per Capita of Europen countries

* Filter observations of European countries
* Group by year
* Summarize the mean gdpPercap
* Save the new data frame as `gdp_Europe`

Answers:
```{r}

```


## Combine data sets

```{r}
df1 <- data.frame(ID = 1:3,                      # Create first example data frame
                  X1 = c("a1", "a2","a3"),
                  stringsAsFactors = FALSE)
df2 <- data.frame(ID = 2:4,                      # Create second example data frame
                  X2 = c("b1", "b2","b3"),
                  stringsAsFactors = FALSE)
df1
df2
```


Left_join: join matching rows from df2 to df1.
```{r}
left_join(df1, df2, by = "ID")
```

Right_join: join matching rows from df1 to df2
```{r}
right_join(df1, df2, by = "ID")
```

Inner_join: retain only rows in both sets
```{r}
inner_join(df1, df2, by = "ID")
```

Full_join: retain all values, all rows
```{r}
full_join(df1, df2, by = "ID")
```


## Working with missing data


### Missing values

NA denotes missing values: is.na() is used to test objects if they are NA
```{r}
x <- c(1, 2, NA, 5, 3)
is.na(x)
```


### Removing NA values

A common task is to remove missing values (NAs)
```{r}
x <- c(1, 2, NA, 4, NA, 5)
```

Exclude vector with all the missing values
```{r}
x[!is.na(x)]
```

If you want to filter out all the missing values for a given variable from a df
```{r}
#df %>% 
  #filter(!is.na(variable))
```

Example of excluding missing values in a df
```{r}
# Create a df with missing values from a previous example
df <- full_join(df1, df2, by = "ID")

# Exclude observations when X1 is NA
df %>% 
  filter(!is.na(X1))
```


What if there are multiple things and you want to take the subset with no missing values?
```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", NA, "b", "d", NA, "f")
```

Leave values that are not missing for both x and y
```{r}
good <- complete.cases(x, y)
x[good]
y[good]
```

Using `complete.case` to filter out NA values from a df
```{r}
df %>% 
  filter(complete.cases(X1,X2))
```


## Making tables look nice in a Markdown file

Use `datatable` function provided by `DT` package
```{r}
#install.packages("DT")
library(DT)
datatable(countries)
```


## Exporting dataframes

Save the dataframe as a csv file
```{r}
write.csv(countries, file="countries.csv")
```


## Sources

* https://statisticsglobe.com/r-dplyr-join-inner-left-right-full-semi-anti 
* https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
